Boosting Higgs Pair Production in the $b\bar{b}b\bar{b}$ Final State with Multivariate Techniques

### 2015-12-07 paper/sec-analysis.tex
more work on analysis section

In section we describe our analysis strategy, and in particular
the classification of each individual event into
different categories according to its topology.
%
First of all, we discuss the settings and nomenclature
for jet clustering, and the strategy for $b$-tagging.
%
Then we discuss the categorisation of events between different
topologies and how to prioritise among them.
%
Finally, we motivate our choice of analysis cuts by comparing signal and background
for representative kinematic distributions.

@@

After the  parton shower, final state particles
are clustered using the
jet reconstruction algorithms
obtained from
{\tt FastJet}~\cite{Cacciari:2011ma,Cacciari:2005hq},
{\tt v3.1.0}.
%
We will use different types of jet definitions:
\begin{itemize}
\item {\it Small-$R$ jets}.

@@

  These jets are also constructed with the
  anti-$k_T$ clustering algorithm, now using a $R=1.0$ radius.
  %
  Large-$R$ jets are required to have
  $p_T \ge 200$~GeV and lie in a pseudo-rapidity region of
  $|\eta|<2.0$.
  %
  The more restrictive range  in pseudo-rapidity
  as compared to the small-$R$ jets
  is motivated by mimicking the  experimental requirements
  in ATLAS and CMS
  related to the track-jet based calibration~\cite{Aad:2014bia,ATLAS:2012kla}.

  In addition to the basic $p_T$ and $\eta$
  acceptance requirements, large-$R$ jets should also
  satisfy the  BDRS mass-drop tagger (MDT)(~\cite{Butterworth:2008iy}
  conditions.
  %
  We use the {\tt FastJet} default
  parameters of  $\mu = 0.67$ and $y_{\textrm{cut}}= 0.09$.
  %
  Before applying the MDT, the large-$R$ jet
  constituents are reclustered with the Cambridge/Aachen (C/A)
  algorithm~\cite{Dokshitzer:1997in,Wobisch:1998wt}
  with $R=1.2$.

@@

It is clear that
optimisation of the $b$-tagging capabilities of the
LHC experiments is an essential ingredient to improve
the prospects for the observation of Higgs pair production in the
$b\bar{b}b\bar{b}$ final state.
%
In this work, we adopt
a $b$-tagging strategy along the lines
of current ATLAS performance~\cite{Aad:2013gja,Aad:2015ydr},
though differences with respect to
the corresponding CMS strategy~\cite{Khachatryan:2011wq,Chatrchyan:2012jua}
should not affect qualitatively our results.
%
For each type of jet definition used, a different
$b$-tagging strategy is adopted:

@@

  If no $b$-quarks are found among the constituents
  of this jet, it can be still be tagged as a $b$-jet with
  a mistag rate of $f_l$, unless a charm quark is present instead,
  and in this case the mistag rate is $f_c$.
  %
  Only jets which contain at least one (light or charm)
  constituent
  with $p_T \ge 15$ GeV will lead to fake $b$-tags.

  We only attempt to $b$-tag only the four (two) hardest small-$R$ jets
  in the resolved (intermediate) category.
  %
  Trying to $b$-tag all the
  small-$R$ jets that satisfy the acceptance cuts worsens the
  overall performance
  due to combinatorics.

@@

    Large-$R$ jets are $b$-tagged by
    ghost-associating anti-$k_T$ $R=0.3$
    subjets to the original large-$R$
    jets~\cite{Cacciari:2007fd,Aad:2013gja,
      ATLAS-CONF-2014-004,Aad:2015uka}.
    %
    A large-$R$ jet will considered $b$-tagged if both
    the leading and subleading AK03 subjets, where the ordering
    is done in the subjet $p_T$, are both individually $b$-tagged,
    with the same criteria as the small-$R$ jets.

    As in the case
    of small-$R$ jets, we only attempt to $b$-tag the two leading subjets,
    else one finds a degradation of the
    signal significance due to combinatorics.

    %
    Therefore, a large-$R$ jet where the two leading
    subjets have at least one $b$-quark will be tagged
    with probability $f_b^2$, if only one of the two leading
    subjets has a $b$-quark, then the tagging probability is
    $2f_bf_l$, and if none of the two subjets include $b$-quarks,
    the mistag rate will be $f_l^2$.

@@

The basic idea of the present analysis is similar to the
scale-invariant resonance tagging introduced
in Ref.~\cite{Gouzevitch:2013qca}.
%
Rather than restricting to a specific event topology,
we aim to consistently combine the information from
the three possible event topologies: boosted, intermediate and
resolved.
%
The optimal cuts for each category will be determined
separately, and then
a exclusive categorisation is defined starting from the category
with the highest signal significance.
%
An advantage of our approach is that the analysis strategy
is robust under variations of
the underlying production model of Higgs pairs is modified,
for instance in the case of
BSM dynamics, which can substantially increase
the degree of boost of the final state.
%
These three categories are defined as follows:
\begin{itemize}
\item {\it Boosted category}.

  An event which
  contains at least two large-$R$ jets, with the two leading
 being $b$-tagged.
 %
 Each of these two large-$R$ $b$-tagged jets are is this
 a candidate
 to contain the decay products of a Higgs boson.

@@

  An event with only one large-$R$ $b$-tagged jet, which
  is assigned to be the leading Higgs candidate.
  %
  In addition, we require at least two small-$R$ $b$-tagged jets,
  which must be separated with respect to the large-$R$ jet
  by $\Delta R\ge 1.0$.
  %

  The subleading Higgs boson candidate is then reconstructed
  by selecting the two $b$-tagged small-$R$ jets that minimize the difference
  between the invariant mass of the large-$R$ jet
  with the invariant mass of the dijet obtained
  from the sum of the two small-$R$ jets.

@@

Once a Higgs boson candidate has been identified,
its invariant mass is required to lie within a fixed window
around the nominal Higgs boson mass.
%
In particular we require the condition
\be
\label{higgsmasswindow}
|m_{h,j} - 125| < 40~{\rm GeV} \, ,\, j=1,2 \, ,
\ee
where $m_{h,j}$ is the invariant mass of each of the two reconstructed  Higgs candidates.
%
This cut is substantially looser as the corresponding
cut used in the typical ATLAS and CMS $h\to b\bar{b}$
analyses~\cite{Aad:2012gxa,Chatrchyan:2013zna}: the motivation
for this choice is that the MVA will take care of the optimisation of these
and other cuts that are not explicitly performed.
%
Only events where the two Higgs candidates satisfy
Eq.~(\ref{higgsmasswindow}) are classified as signal events.
%

@@

 This is achieved as follows.
 %
First of all we perform an inclusive analysis, and optimise the
signal significance
$S/\sqrt{B}$ in each of the three categories separately, including
the MVA.
%
We find that the category with higher significance turns out to be the boosted,
followed by the intermediate and the resolved ones, the two latter
with similar significance.
%
Therefore for the final analysis we turn to an exclusive approach:
if an event satisfies the boosted requirements, it is kept for
these category, else we check if it suits the intermediate
requirements, and failing also this, the event is attempted to
be classified in the
resolved category.
%
The resulting exclusive event samples are then separately processes
through the MVA.
%
This allows a consistent combination
of the significance of the three event categories.

\subsection{Optimisation of kinematical cuts}

We now motivate the values of
kinematical cuts applied to the different categories.
%
These cuts have been separately
optimised in an inclusive analysis, since it is only
once we have determined
the dominant category that the analysis is made exclusive.
%
Here we show the various kinematic distributions 
before the corresponding cut has been applied.
%
For illustration purposes,
only a representative subset of
all available kinematic distributions will be shown here.
%
For simplicity,
we restrict to the discussion of the
kinematic distributions of the boosted
and resolved categories.

First of all,
in Fig.~\ref{fig:cutplots1} we show
the $p_T$ distribution of the 
of the
  leading and subleading large-$R$ jets in the boosted category, comparing
  the shapes of the distribution in signal and background events.
  %
  Distributions have been normalized to their total integral.
  %
  We see that the background distribution
  falls off rather steeply than the signal at large $p_T$
  values, therefore
  providing evidence that exploiting the boosted region will be
  advantageous to enhance the signal over background ratio.
  %
  This is true both for the leading and the subleading
  large-$R$ jets.
  %
  These distributions justify our choice of $p_T \ge 200$ GeV
  for the large-$R$ jets in the boosted category.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{figure}[t]
\begin{center}
 \includegraphics[width=0.48\textwidth]{plots/pt_H0_boost_C1.pdf}
 \includegraphics[width=0.48\textwidth]{plots/pt_H1_boost_C1.pdf}
\caption{\small  The $p_T$ distribution of the
  leading (left plot) and
  subleading (right plot) large-$R$ jets in the boosted category.
  %
  We comparing
  the shapes of the distribution in signal and background events,
  where both
  distributions have been normalized to their total integral.
}
\label{fig:cutplots1}
\end{center}
\end{figure}
%%%%%%%%%%%%%%%%%%%%%%%

In the boosted category we require the two
leading subjets of the large-$R$ jet to be relatively
hard, in particular that should satisfy $p_T \ge $ 50 GeV.
%
To motivate this cut, in Fig.~\ref{fig:cutplots22}
we show the distribution in $p_T$ of the leading
and subleading sub-jets in the leading large-$R$ jet in events
of the boosted category.
%
It is clear from the comparison that the subjet $p_T$ spectrum is
relatively harder in the signal with respect to the background,
motivating our choice of subjet $p_T$ cut.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{figure}[t]
\begin{center}
  \includegraphics[width=0.48\textwidth]{plots/m_HH_res_C1.pdf}
  \includegraphics[width=0.48\textwidth]{plots/m_HH_boost_C1.pdf}
  \caption{\small Left plot: the invariant mass distribution of the Higgs
    pair candidates, $m_{hh}$, comparing signal and background events,
    in the resolved category.
    %
    Right plot: same for the boosted category.
}
\label{fig:mhh}
\end{center}
\end{figure}
%%%%%%%%%%%%%%%%%%%%%%%

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{figure}[t]
\begin{center}
  \includegraphics[width=0.48\textwidth]{plots/pt_HH_res_C1.pdf}
  \includegraphics[width=0.48\textwidth]{plots/pt_HH_boost_C1.pdf}
  \caption{\small Same as Fig.~\ref{fig:mhh} for the transverse momentum
    distribution of the di-Higgs system $p_T^{hh}$.
}
\label{fig:pthh}
\end{center}
\end{figure}
%%%%%%%%%%%%%%%%%%%%%%%

\subsection{PU subtraction with {\tt SoftKiller}}

>>>>

In section we describe our analysis strategy.
%
First of all, we discuss the settings
for jet clustering  and the strategy for $b$-tagging.
%
Then we discuss the categorisation of events in different
topologies and how to prioritise among them.
%
We motivate our choice of analysis cuts by comparing signal and background
for representative kinematical distributions.
%
Finally, we describe the simulation of PU and validate
the PU subtraction strategy that we adopt.

@@

After the  parton shower, final state particles
are clustered using the
jet reconstruction algorithms
obtained from
{\tt FastJet}~\cite{Cacciari:2011ma,Cacciari:2005hq},
{\tt v3.1.0}.
%
We use different types of jet definitions:
\begin{itemize}
\item {\it Small-$R$ jets}.

@@

  These jets are also constructed with the
  anti-$k_T$ clustering algorithm, now using a $R=1.2$ radius.
  %
  Large-$R$ jets are required to have
  $p_T \ge 200$~GeV and lie in a pseudo-rapidity region of
  $|\eta|<2.0$.
  %
  The more restrictive range  in pseudo-rapidity
  as compared to the small-$R$ jets
  is motivated by mimicking the  experimental requirements
  in ATLAS and CMS
  related to the track-jet based calibration~\cite{Aad:2014bia,ATLAS:2012kla}.

  In addition to the basic $p_T$ and $\eta$
  acceptance requirements, large-$R$ jets should also
  satisfy the  BDRS mass-drop tagger (MDT)~\cite{Butterworth:2008iy}
  conditions.
  %
  We use the {\tt FastJet} default
  parameters of  $\mu = 0.67$ and $y_{\textrm{cut}}= 0.09$.
  %
  Before applying the MDT, the large-$R$ jet
  constituents are reclustered with the Cambridge/Aachen (C/A)
  algorithm~\cite{Dokshitzer:1997in,Wobisch:1998wt}
  with $R=1.2$.

@@

In this analysis we adopt
a $b$-tagging strategy along the lines
of current ATLAS performance~\cite{Aad:2013gja,Aad:2015ydr},
though differences with respect to
the corresponding CMS case~\cite{Khachatryan:2011wq,Chatrchyan:2012jua}
do not modify qualitatively our results.
%
For each type of jet definition used, a different
$b$-tagging strategy is adopted:

@@

  If no $b$-quarks are found among the constituents
  of this jet, it can be still be tagged as a $b$-jet with
  a mistag rate of $f_l$, unless a charm quark is present instead,
  and in this case the mistag rate is $f_c$.
  %
  Only jets that contain at least one (light or charm)
  constituent
  with $p_T \ge 15$ GeV can induce a fake $b$-tag.

  We attempt to $b$-tag only the four (two) hardest small-$R$ jets
  in the resolved (intermediate) category.
  %
  Trying to $b$-tag all the
  small-$R$ jets that satisfy the acceptance cuts worsens the
  overall performance
  due to combinatorics.

@@

    Large-$R$ jets are $b$-tagged by
    ghost-associating anti-$k_T$ $R=0.3$ (AKT03)
    subjets to the original large-$R$
    jets~\cite{Cacciari:2007fd,Aad:2013gja,
      ATLAS-CONF-2014-004,Aad:2015uka}.
    %
    A large-$R$ jet will considered $b$-tagged if both
    the leading and subleading AK03 subjets, where the ordering
    is done in the subjet $p_T$, are both individually $b$-tagged,
    with the same criteria as the small-$R$ jets.
    %
     Therefore, a large-$R$ jet where the two leading
    subjets have at least one $b$-quark will be tagged
    with probability $f_b^2$.

    As in the case
    of small-$R$ jets, we only attempt to $b$-tag the two leading subjets,
    else one finds a degradation of the
    signal significance due to combinatorics.
    %
    The simulation of the fakes from light and charm jets
    is the same as in the small-$R$ jets.

@@

The basic idea of the present analysis is similar to the
scale-invariant resonance tagging introduced
in Ref.~\cite{Gouzevitch:2013qca}.
%
Rather than restricting to a specific event topology,
we aim to consistently combine the information from
the three possible topologies: boosted, intermediate and
resolved.
%
The optimal cuts for each category will be determined
separately, and then
a exclusive categorisation is defined starting from the category
with the highest signal significance.
%
An advantage of our approach
the robustness under variations of
the underlying production model of Higgs pairs is modified,
for instance in the case of
BSM dynamics, which can substantially increase
the degree of boost of the final state.

These three categories are defined as follows:
\begin{itemize}
\item {\it Boosted category}.

  An event which
  contains at least two large-$R$ jets, with the two leading jets
 being $b$-tagged.
 %
 Each of these two large-$R$, $b$-tagged, jets are 
 thus candidates
 to contain the decay products of a Higgs boson.

@@

  An event with only one large-$R$, $b$-tagged, jet, which
  is assigned to be the leading Higgs candidate.
  %
  In addition, we require at least two small-$R$, $b$-tagged, jets,
  which must be separated with respect to the large-$R$ jet
  by an angular distance of $\Delta R\ge 1.2$.
  %

  The subleading Higgs boson candidate is reconstructed
  by selecting the two $b$-tagged small-$R$ jets that minimize the difference
  between the invariant mass of the large-$R$ jet
  with that of the dijet obtained
  from the sum of the two small-$R$ jets.

@@

Once a Higgs boson candidate has been identified,
its invariant mass is required to lie within a fixed window
around the nominal Higgs boson mass~\cite{Aad:2015zhl}.
%
In particular we require the condition
\be
\label{higgsmasswindow}
|m_{h,j} - 125| < 40~{\rm GeV} \, ,\, j=1,2 \, ,
\ee
where $m_{h,j}$ is the invariant mass of each of the two reconstructed  Higgs candidates.
%
This cut is substantially looser as the corresponding
cut used in the typical ATLAS and CMS $h\to b\bar{b}$
analyses~\cite{Aad:2012gxa,Chatrchyan:2013zna}: the motivation
for this choice is that the MVA will take care of the optimisation of this,
as well as of other cuts that are not explicitly performed.
%
Only events where the two Higgs candidates satisfy
Eq.~(\ref{higgsmasswindow}) are classified as signal events.
%

@@

 This has been achieved as follows.
 %
First of all we perform an inclusive analysis, and optimise the
signal significance
$S/\sqrt{B}$ in each of the three categories separately, including
the MVA.
%
We find that the category with higher significance turns out to be the boosted,
followed by the intermediate and the resolved ones, the two latter
with similar significance.
%
Therefore, for the final analysis, we turn to an exclusive approach:
if an event satisfies the boosted requirements, it is assigned to
this category, else we check if it suits the intermediate
requirements, and failing also this, the event is attempted to
be classified in the
resolved category.
%
The resulting exclusive event samples are then separately processed
through the MVA, allowing a consistent combination
of the significance of the three event categories.

\subsection{Motivation for kinematical cuts}

We now motivate the values of
kinematical cuts applied to the different categories, 
comparing representative kinematic distributions between
signal and background events.
%
First to all we present results without PU, and in the next
subsection we discuss what is the impact of PU
on the description of the various kinematical
distributions.
%
In the following, all
distributions will be normalized to their total integral.

%
To begin with,
in Fig.~\ref{fig:cutplots1} we show
the $p_T$ distribution of the 
of the
  leading and subleading large-$R$ jets in the boosted category, comparing
  the shapes of the distribution in signal and background events.
  %
  We see that the background distribution
  falls off rather steeply than the signal at large $p_T$
  for the two Higgs candidates.
  %
  On the other hand, the cut in $p_T$ cannot be too strong to avoid
  a substantial degradation of signal selection efficiency,
  specially taking into account the subleading large-$R$ jet.
  %
  This comparison justifies the cut of $p_T \ge 200$ GeV
  for the large-$R$ jets that we impose in the boosted category.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{figure}[t]
\begin{center}
 \includegraphics[width=0.48\textwidth]{plots/pt_H0_bst_C1d_noPU.pdf}
 \includegraphics[width=0.48\textwidth]{plots/pt_H1_bst_C1d_noPU.pdf}
\caption{\small  Comparison of the $p_T$ distributions of the
  leading (left plot) and
  subleading (right plot) large-$R$ jets in the boosted category,
  in in signal and background events.
  %
  Distributions have been normalized to unity.
}
\label{fig:cutplots1}
\end{center}
\end{figure}
%%%%%%%%%%%%%%%%%%%%%%%

Also in the boosted category we require the two
leading subjets of the large-$R$ jet to be relatively
hard, in particular that should satisfy $p_T \ge $ 50 GeV.
%
To motivate this cut, in Fig.~\ref{fig:cutplots22}
we show the distribution in $p_T$ of the leading
and subleading sub-jets in the leading large-$R$ jet in events
of the boosted category.
%
It is clear from the comparison that the subjet $p_T$ spectrum is
relatively harder in the signal with respect to the background,
motivating our choice of subjet $p_T$ cut.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{figure}[t]
\begin{center}
  \includegraphics[width=0.48\textwidth]{plots/m_HH_C2_res_noPU.pdf}
  \includegraphics[width=0.48\textwidth]{plots/m_HH_C2_bst_noPU.pdf}
  \caption{\small Left plot: the invariant mass distribution of the Higgs
    pair candidates, $m_{hh}$, comparing signal and background events,
    in the resolved category.
    %
    Right plot: same for the boosted category.
}
\label{fig:mhh}
\end{center}
\end{figure}
%%%%%%%%%%%%%%%%%%%%%%%

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{figure}[t]
\begin{center}
  \includegraphics[width=0.48\textwidth]{plots/pt_HH_C2_res_noPU.pdf}
  \includegraphics[width=0.48\textwidth]{plots/pt_HH_C2_bst_noPU.pdf}
  \caption{\small Same as Fig.~\ref{fig:mhh} for the transverse momentum
    distribution of the di-Higgs system $p_T^{hh}$.
}
\label{fig:pthh}
\end{center}
\end{figure}
%%%%%%%%%%%%%%%%%%%%%%%

\subsection{PU subtraction with {\tt SoftKiller}}
\label{sec:pileup}
